CPU-Scheduling

### 2024-11-11 Archive/Handout/article.tex
read README 2024-11-11

% TODO: Explain why a CPU can only do one operation at the time
We all use computers.
The modern person couldn't do half of the things that is expected from him without an instance of Windows, MacOs or any other Operating System.
We surf the web, write emails, documents and have a video call at the same time.
The question is: "How does CPU achieve all of this?" The things listed are only the parts of the OS that we come in contact with on a daily basis.
What about all the other things hidden under the hood? If it comes down to doing one thing than it's fine.
The problem arises then we have multiple tasks: The CPU can only do one thing at a time and yet we all multitask on our machines.
This Black Box effect of taking multiple jobs and working on them in a proper order is what I will disect in this "Maturarbeit".

@@

\begin{defi}[Turnaround Time]{dfn:turnaround}
The turnaround time is a performance metric. It measures how long a task took to be completed from the time it arrived.
It is calculated as follows:
$$T_{Turnaround} = T_{Completion} - T_{Arrival}$$
\vspace{1em}
\end{defi}

@@

% TODO: MAYBE HERE THE METRICS PART?

@@

There are two notable changes in this code snippet when compared to FIFO. One of them is relevant.
First, the less exiting part is that I actually took my time to write out some of the process class definition.
Second and the real difference to FIFO is that we sort the queue according to the burst time\footnote{Burst time is basically the runtime needed until the process finishes. This variable can be only predicted}. 
This means that we will choose the task with the shortest time to completion.
% TODO: AT THIS PART THE KNOWLEDGE OF PROPER SORTING ALGORITHMS IS MISSING: MAYBE A REVERSE BUBBLE SORT? 
% TODO: => Sorting algorithm input?
% TODO: THE LIST SHOULD BE ALREADY SOMEWHAT ORDERED AND THE LAST MEMBER HAS TO BE SORTED
The only problem is when you arrive late. The scanning of products already started and you'll have to wait.
This policy is a so called non-preemptive policy.
"Non-preemptive" just means that once a task will run until it finishes.

@@

If SJF and FIFO are cousins that STCF is the brother.
Here we will make SJF "preemptive", meaning that our cashier can now "save" the state of the previously scanned items. 
Therefore STCF is also called "Preemptive Shortest Job First (PSJF)".
Although the switching costs some time it does not matter here, because one would have to wait until the little boy returns with the ketchup\footnote{If you don't know what I am talking about read Chapter 1.1 (FIFO)}
% TODO: Pretty change perspective switch, should revisit it later
The option of saving a state comes in handy, especially if a program has to wait for I/O.
The state of waiting is usually called "blocked" and the technique of running something else during the task is blocked is called overlapping.
Preemptiveness also makes the computer much more response, because now you can really do certain things "at the same time".
The best part is that we don't even have to change much of the previous code.
\pagebreak
\begin{lstlisting}[language=Python, style=colorEX, label=cd:stcf-python]
  # Simple implementation of STCF in Python

@@

    use_resource(current) # function that will allow "next" to use the CPU (also tracks runtime)
\end{lstlisting}
% TODO: IDK ABOUT THE CODE; IT IS JUST KINDA HACKED TOGETHER; IT IS DEFINITELY NOT SECURE WHATSOEVER! Also should explain it more in words!
% TODO: GENErally code is trash. should do a more RR style, with after every allotment new scheduling
Whenever a new process "joins", the "allocation" function gets run.
In addition to that, whenever the process finishes, or it voluntarily gives up the CPU, the same "allocation" function can be run.

@@

Now, why don't we just scan one item for everyone? 
% TODO: Exlain how that would make it better...
This would mean that, even though slow but progress is being constantly made.
Even if at this point the item scanning doesn't really work anymore\footnote{You can't scan a fraction of a product, or can you?}, let's just make the so called "quanta" (quantum in singular) even smaller.
% TODO: Explain term quantum better
As already mentionned in the previous chapter, the cashier needs some time to "save" the state of the current purchases. 
This effectively mean that after the quantum shrinks to a certain extent, the "context switching"\footnote{Context switch saves the state of a process so that it can be restored later} would get too expensive. 
The goal is to reach a balance, where we have enough responsibility, but the price for it is not too high.
Notice how by treating the processes more "fairly" by don't caring about their overall burst time, we get rid of a variable, which can only be predicted and replace it with another one, which is more abstract and global.
\pagebreak
\begin{lstlisting}[language=Python, style=colorEX, label=cd:rr-python]
  # Simple implementation of RR in Python

@@

The most basic principle of MLFQ are the so called "Priority Queues". These are priorities that a program can belong into.
% TODO: SIMILAR TO PREVIOUS QUEUES??? MAKE A CONNECTINjjjjkk
They can change over time and these changes reflect the action/state of a program.
The changing of the priority is done as followed: every task has a certain amount of "allotment" for each priority.
If this CPU time is used up, than the process will be moved down once and recieve a new set of allotment.

@@

As you can see processes at a higher priority are preferred over processes lower down. 
This effectively means that once process 2 joins the comptetition it can monopolize the CPU until they end up on the same level.
Once they are there Round Robin is used to give them both a fair share of allotment.
With each process starting in the highest queue, this means that the shorter tasks are generally preferred, because they can finish quickly before they end up on a lower queue. Therefore the system becomes more responsive.
Keep in mind that if a process gives up it's CPU before the allotment is used up (for example if they wait for I/O), than they will "stay" in the same priority and can use up the rest of the CPU time, before getting demoted.
% TODO: INSERT WATING TIEM EXAMPLE HERE
Therefore not the actual time spent there matters, but the time spent using the resource.
As you might have to already noticed poor process 1 didn't get any runtime while process 2 was getting their priority prioperly adjusted. Note: for simplicity there are now only 4 queues.

@@

\noindent So as "OSTEP REFERENCE HERE" would summarize it: % TODO: DO I EVEN NEED THIS?
\begin{enumerate}
  \item If Priority (A) $>$ Priority (B) $\Rightarrow$ A runs \& B doesn't
  \item If Priority (A) = Priority (B) $\Rightarrow$ A \& B run in RR
  \item When a job enters the system, it is placed at the highest priority
  \item Once a job uses up its time allotment at a given level, its priority is reduced
  \item After some period S, move all the jobs in the system to the topmost queue
\end{enumerate}

>>>>

We all use computers.
The modern person couldn't do half of the things that is expected from him without an instance of Windows, MacOs or any other Operating System.
We surf the web, write emails, documents and have a video call at the same time.
The question is: "How does CPU achieve all of this?" The things listed are only the parts of the OS that we come in contact with on a daily basis.
What about all the other things hidden under the hood?
If simplified, a simple CPU is like a simple production line. It does one things at a time.
Therefore a problem arises then we have multiple tasks: The CPU can only do one thing at a time\footnote{When talking about a single-core CPU for simplification} and yet we all multitask on our machines.
This Black Box effect of taking multiple jobs and working on them in a proper order is what I will disect in this "Maturarbeit".

@@

% TODO: WAY TOO DEEP, BECAUSE WILL NEVER REALLY USE IT AGAIN THAT DEEPLY (EXCEPT IF A PROPER EXPLANATION IS DONE FOR EACH POLICY)
\begin{defi}[Turnaround Time]{dfn:turnaround}
The turnaround time is a performance metric. It measures how long a task took to be completed from the time it arrived.
It is calculated as follows:
$$T_{Turnaround} = T_{Completion} - T_{Arrival}$$
\vspace{1em}
\end{defi}

@@

@@

There are two notable changes in this code snippet when compared to FIFO. One of them is relevant.
First, the less exiting part is that I actually took my time to write out some of the process class definition.
Second and the real difference to FIFO is that we sort the queue according to the burst time\footnote{Burst time is basically the runtime needed until the process finishes. This variable can be only predicted}. 
This means that we will choose the task with the shortest time to completion.
At this point one could go into which sort algorithm to use and which one is the most efficient, however that is not something I will touch upon here.
The reason is that I don't think that it is necessary to understand those in order to grasp the concept of SJF.

Back to our story, now you only have a problem when you arrive late. The scanning of products has already started and you'll have to wait.
This policy is a so called non-preemptive policy.
"Non-preemptive" just means that once a task will run until it finishes.

@@

If SJF and FIFO are cousins than STCF is the brother.
To fix the issue from before, we'll make SJF "preemptive", meaning that our cashier can now "save" the state of the previously scanned items. 
Therefore STCF is also called "Preemptive Shortest Job First (PSJF)".
The option of saving a state comes in handy, especially if a program has to wait for I/O.
Although the switching costs some time, this can be disregarded, because one would have to wait until the little boy returns with the ketchup\footnote{If you don't know what I am talking about read Chapter 1.1 (FIFO)} anyways.
The state of waiting is usually called "blocked" and the technique of running something else during the task is blocked is called overlapping.
% TODO: explain blocked and overlapping with the cashier in a better way
Preemptiveness also makes the computer much more responsive, because now you can really do certain things "at the same time".
The best part is that we don't even have to change much of the previous code.
\pagebreak
\begin{lstlisting}[language=Python, style=colorEX, label=cd:stcf-python]
  # Simple implementation of STCF in Python

@@

    use_resource(current) # function that will allow "next" to use the CPU (also tracks runtime)
\end{lstlisting}
Whenever a new process "joins", the "allocation" function gets run.
In addition to that, whenever the process finishes, or it voluntarily gives up the CPU, the same "allocation" function can be run.

@@

Now, why don't we just scan one item for everyone? 
This could mean that every one sees progress and that same sized jobs finish at almost the same time
Even if at this point the item scanning explanation doesn't really work anymore\footnote{You can't scan a fraction of a product, or can you?}, let's just make the so called "quanta" (quantum in singular) even smaller.
% TODO: Explain term quantum better
As already mentionned in the previous chapter, the cashier needs some time to "save" the state of the current purchases. 
This effectively mean that after the quantum shrinks to a certain extent, the "context switching"\footnote{Context switch saves the state of a process so that it can be restored later} would get too expensive. 
The goal is to reach a balance, where we have enough responsibility, but the price for it is not too high.
Notice how by treating the processes more "fairly" by don't caring about their overall burst time, we get rid of a variable, which can only be predicted and replace it with another one, which is more abstract and global.
\pagebreak
\begin{lstlisting}[language=Python, style=colorEX, label=cd:rr-python]
  # Simple implementation of RR in Python

@@

The most basic principle of MLFQ are the so called "Priority Queues". These are priorities that a program can belong into.
They can change over time and these changes reflect the action/state of a program.
The changing of the priority is done as followed: every task has a certain amount of "allotment" for each priority.
If this CPU time is used up, than the process will be moved down once and recieve a new set of allotment.
This way you could have a fast lane checkout, where you can only scan a few items.
A separation of small and bulk purchases would occur making the experience overall faster.
There is still only one cashier though, so there are some pitfall that one must avoid.

@@

As you can see processes at a higher priority are preferred over processes lower down. 
This effectively means that once process 2 joins the comptetition it can monopolize the CPU until they end up on the same level.
Once they are there Round Robin is used to give them both a fair share of allotment.
With each process starting in the highest queue, this means that the shorter tasks are generally preferred, because they can finish quickly before they end up on a lower queue. Therefore the system becomes more responsive.
Keep in mind that if a process gives up it's CPU before the allotment is used up (for example if they wait for I/O), than they will "stay" in the same priority and can use up the rest of the CPU time, before getting demoted.
% TODO: INSERT WATING TIEM EXAMPLE HERE meaning a sketch
Therefore not the actual time spent there matters, but the time spent using the resource.
As you might have to already noticed poor process 1 didn't get any runtime while process 2 was getting their priority prioperly adjusted. Note: for simplicity there are now only 4 queues.

@@

\noindent So as "OSTEP REFERENCE HERE" would summarize it: % TODO: find and insert reference
\begin{enumerate}
  \item If Priority (A) $>$ Priority (B) $\Rightarrow$ A runs \& B doesn't
  \item If Priority (A) = Priority (B) $\Rightarrow$ A \& B run in RR
  \item When a job enters the system, it is placed at the highest priority
  \item Once a job uses up its time allotment at a given level, its priority is reduced
  \item After some period S, move all the jobs in the system to the topmost queue
\end{enumerate}
